Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments

### 2021-08-30 doc/paper.Rmd
adresses #76

Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwoodEffectsClimateChange2015; @williams2015], and that the effects of climate change could be more pronounced when habitat is fragmented [@oliver2015; @holyoak2016].
<!--# explain why R1.2 --> While the demographic consequences of climate change for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the responses of these species to climatic variability [@paniw2021].
Climate models predict increases in the areas of the Amazon effected by severe drought and unusual wetness by the year 2100 [@duffyProjectionsFutureMeteorological2015].
Tropical plants may be particularly sensitive to climate change---they typically have narrow ranges of climatic tolerance [@feeley2012], and recent results suggest increases in the frequency and severity of extreme precipitation events (both drought and extreme wet) reduce survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
This sensitivity to climatic fluctuations, coupled with evidence that plant growth and survivorship are lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in forest fragments will be especially susceptible to climate change [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015].

>>>>

Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwoodEffectsClimateChange2015; @williams2015].
While the demographic consequences of climate change for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the responses of these species to climatic variability [@paniw2021].
Climate models predict increases in the areas of the Amazon effected by severe drought and unusual wetness by the year 2100 [@duffyProjectionsFutureMeteorological2015].
Tropical plants may be particularly sensitive to climate change---they typically have narrow ranges of climatic tolerance [@feeley2012], and recent results suggest increases in the frequency and severity of extreme precipitation events (both drought and extreme wet) reduce survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
This sensitivity to climatic fluctuations, coupled with evidence that plant growth and survivorship are lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in forest fragments will be especially susceptible to climate change [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015].

The simultaneous pressures of climate change and habitat fragmentation may also result in worse than additive impacts on demography [@oliver2015; @holyoak2016].
For example, if fragments have a reduced capacity to buffer changes in microclimate [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999] or if fragment connectivity is climate dependent [@honnayPossibleEffectsHabitat2002], populations experiencing habitat fragmentation will fare worse under climate change.
